Tree height and hydraulic traits shape growth responses across droughts in a temperate broadleaf forest

### 2020-07-17 manuscript/response_to_reviews.Rmd

Dear Editor:

... Major changes included the following:

* At the request of 3 reviewers (R1,R3,R4), we tried an alternate metric of drought resistance (using an ARIMA model to project what growth would have been in drought year based on the trend over the past 10 years). Results were similar, and we now present that as a secondary model (mostly in SI).

* We dropped crown position as a factor in the statistical models of drought resistance, and instead include it as part of an enhanced focus on the differences in microenviroment experienced by trees of different height (now in Table 1; Fig. 2). This change was prompted by comments by two reviewers (**R1, R3**) that canopy position could not be reliably projected back to 1966 or disentagled from height.

@@

L266-267: Don’t think you can test the effects of crown dominance going back this far. How can you be sure a tree was dominant/suppressed back in 1966?

L287: specify how they were related to drought. Positive? Negative?

Tables and Figures: there are a large number of tables. I wonder if Table 2 and 3 would best fit in supporting information (with key information integrated directly into the text). I also feel that it would be nice to represent some of the key results (e.g., Rt and its relationship to height and leaf hydraulic traits) in a figure.

**We have moved two tables (former tables 4 and 5 to the SI). The best models are now visualized in a figure (Fig. 4). We have also added a figure showing drought response by species.**

### 2020-07-17 manuscript/response_to_reviews.tex

\textasciitilde\textasciitilde\textasciitilde\textasciitilde{}

Dear Editor:

\ldots{} Major changes included the following:

\begin{itemize}
\item
  At the request of 3 reviewers (R1,R3,R4), we tried an alternate metric
  of drought resistance (using an ARIMA model to project what growth
  would have been in drought year based on the trend over the past 10
  years). Results were similar, and we now present that as a secondary
  model (mostly in SI).
\item
  We dropped crown position as a factor in the statistical models of
  drought resistance, and instead include it as part of an enhanced
  focus on the differences in microenviroment experienced by trees of
  different height (now in Table 1; Fig. 2). This change was prompted by
  comments by two reviewers (\textbf{R1, R3}) that canopy position could
  not be reliably projected back to 1966 or disentagled from height.
\item
  We added two figures: one showing Rt differences across species, and
  one visualizing the model results (the latter is not yet right, as
  gg-plot is being difficult, but will eventually match the model
  results.)
\item
  We moved former tables 4 and 5 to the SI.
\end{itemize}

@@

L266-267: Don't think you can test the effects of crown dominance going
back this far. How can you be sure a tree was dominant/suppressed back
in 1966?

L287: specify how they were related to drought. Positive? Negative?

Tables and Figures: there are a large number of tables. I wonder if
Table 2 and 3 would best fit in supporting information (with key
information integrated directly into the text). I also feel that it
would be nice to represent some of the key results (e.g., Rt and its
relationship to height and leaf hydraulic traits) in a figure.

\textbf{We have moved two tables (former tables 4 and 5 to the SI). The
best models are now visualized in a figure (Fig. 4). We have also added
a figure showing drought response by species.}

@@

\textbf{To clarify that this shouldn't be an issue, we have added the
following sentence (methods section): ``We note that drought was
probably not a cause of mortality for these trees, as monthly May-Aug
\(PDSI\) did not drop below -1.75 in these years or the three years
prior (2013-2017), and that trees cored dead displayed similar climate
sensitivity to trees cored live (Helcoski \emph{et al.}, 2019).''}

L160: you mean a log-log regression, right?

L181: N=3 individuals per species is really quite low. In addition, by
sampling up to 8 m height you would probably be sampling less exposed
branches from taller trees, which could confound your results.

\emph{from Lawren: ``Re the trait values, I believe we would see species
differences, even with n=3, looking at the means and SD presented.
Suggest to run an ANOVA on each trait (and/or a MANOVA on all traits)
and put the significance values in a new row in Table 3. (BTW also
suggest limiting values in Table 3 to 3 sig figs, and presenting SE in
that Table with the means rather than SD).''}

### 2020-07-17 manuscript/climate_sensitivity_draft.Rmd
minor work, mostly on response to reviews

Many studies have shown that within species, large trees tend to be more affected by drought. Greater growth reductions for larger trees was first shown on a global scale by @bennett_larger_2015, and subsequent studies have reinforced this finding (*e.g.*, @stovall_tree_2019; @hacket-pain_consistent_2016). It has yet to be resolved which of several potential underlying mechanisms most strongly shape size trends in drought response. First, tree height may be a primary driver. Taller trees face the biophysical challenge of lifting water greater distances against the effects of gravity and friction [@mcdowell_relationships_2011; @mcdowell_darcys_2015; @ryan_hydraulic_2006; @couvreur_water_2018]. Vertical gradients in stem and leaf traits--including smaller and thicker leaves (higher leaf mass per area, LMA), greater resistance to hydraulic dysfunction (*i.e.*, more negative water potential at 50% loss of hydraulic conductivity, more negative P50), and lower hydraulic conductivity at greater heights [@couvreur_water_2018; @koike_leaf_2001; @mcdowell_relationships_2011]--enable trees to become tall [@couvreur_water_2018]. Indeed, tall trees require xylem of greater hydraulic efficiency, such that xylem conduit diameters are wider in the basal portions of taller trees, both within and across species [@olson_plant_2018; @liu_hydraulic_2019], and throughout the conductive systems of angiosperms (*Zak et al. 2010, Olson et al. 2014,2018*). Wider xylem conduits plausibly make large trees more vulnerable to embolism during drought [@olson_plant_2018], and traits conducive to efficient water transport may also lead to poor ability to recover from or re-route water around embolisms [@roskilly_conflicting_2019]. Larger trees may also have lower drought resistance because their crowns tend to occupy more exposed canopy positions, which are associated with higher evaporative demand (**REFS--KAT**)[@kunert_revised_2017]. Subcanopy trees tend to fare better specifically due to the benefits of a buffered environment [@pretzsch_drought_2018]. Counteracting the liabilities of associated with tall height, large trees tend to have larger root systems, which potentially counteracts some of the biophysical challenges they face by allowing greater access to water. Finally, tree size-related responses to drought can be modified by species' traits and their distribution across size classes [@meakem_role_2018; @liu_hydraulic_2019]. Understanding the mechanisms driving the greater relative growth reductions of larger trees during drought will require sorting out the interactive effects of height and associated exposure, root water access, and species' traits.

>>>>

Many studies have shown that within species, large trees tend to be more affected by drought. Greater growth reductions for larger trees was first shown on a global scale by @bennett_larger_2015, and subsequent studies have reinforced this finding (*e.g.*, @stovall_tree_2019; @hacket-pain_consistent_2016). It has yet to be resolved which of several potential underlying mechanisms most strongly shape size trends in drought response. First, tree height may be a primary driver. Taller trees face the biophysical challenge of lifting water greater distances against the effects of gravity and friction [@mcdowell_relationships_2011; @mcdowell_darcys_2015; @ryan_hydraulic_2006; @couvreur_water_2018]. Vertical gradients in stem and leaf traits--including smaller and thicker leaves (higher leaf mass per area, LMA), greater resistance to hydraulic dysfunction (*i.e.*, more negative water potential at 50% loss of hydraulic conductivity, more negative P50), and lower hydraulic conductivity at greater heights [@couvreur_water_2018; @koike_leaf_2001; @mcdowell_relationships_2011]--enable trees to become tall [@couvreur_water_2018]. Indeed, tall trees require xylem of greater hydraulic efficiency, such that xylem conduit diameters are wider in the basal portions of taller trees, both within and across species [@olson_plant_2018; @liu_hydraulic_2019], and throughout the conductive systems of angiosperms (*Zak et al. 2010, Olson et al. 2014,2018*). Wider xylem conduits plausibly make large trees more vulnerable to embolism during drought [@olson_plant_2018], and traits conducive to efficient water transport may also lead to poor ability to recover from or re-route water around embolisms [@roskilly_conflicting_2019]. Larger trees may also have lower drought resistance because their crowns tend to occupy more exposed canopy positions, which are associated with higher evaporative demand [@kunert_revised_2017]. Subcanopy trees tend to fare better specifically due to the benefits of a buffered environment [@pretzsch_drought_2018]. Counteracting the liabilities of associated with tall height, large trees tend to have larger root systems, which potentially counteracts some of the biophysical challenges they face by allowing greater access to water. Finally, tree size-related responses to drought can be modified by species' traits and their distribution across size classes [@meakem_role_2018; @liu_hydraulic_2019]. Understanding the mechanisms driving the greater relative growth reductions of larger trees during drought will require sorting out the interactive effects of height and associated exposure, root water access, and species' traits.
